High-Dimensional Regularized Discriminant Analysis

### 2013-12-29 reports/paper-hdrda.Rnw
Initial version of algorithm added. Moving forward on #11

\usepackage{graphicx, amsmath, amssymb, bm, mathtools, amsthm}
\usepackage{natbib, setspace}

@@

\begin{proposition}\label{proposition:hdrda-W_k}
Let $\bm W_k$ be defined as above, $k = 1, \ldots, K$. Then, $|\bm W_k| = |
\bm \Gamma_k| |\bm Q_k|$ and
	\begin{align}
		\bm W_k^{-1} &= \bm \Gamma_k^{-1} - \alpha_k(1 - \lambda) \bm \Gamma_k^{-1} \bm U_1^{T} \bm X_k^{T} \bm Q_k^{-1} \bm X_k \bm U_1 \bm \Gamma_k^{-1},\label{eq:W_k_inv}
	\end{align}
where $\bm Q_k = \bm I_{n_k} + \alpha_k(1 - \lambda) \bm X_k \bm U_1 
\bm \Gamma_k^{-1} \bm U_1^{T} \bm X_k^{T}$ and $\bm \Gamma_k = \alpha_k \lambda \bm D_q +
\gamma \bm I_q$.
\end{proposition}
\begin{proof}
To calculate $|\bm W_k|$, we apply Theorem 18.1.1 from \cite{Harville:2008wja},
which states that $|\bm A + \bm B \bm T \bm C| = |\bm A| |\bm T| |\bm T^{-1} +
\bm C \bm A^{-1} \bm B|$. Thus, setting $\bm A = \bm \Gamma_k$, $\bm B =
\alpha_k (1 - \lambda) \bm U_1^{T} \bm X_k^{T}$, $\bm T = \bm I_{n_k}$, and $\bm
C = \bm X_k \bm U_1$, we have $|\bm W_k| = |\bm \Gamma_k| |\bm Q_k|$. Also,
\eqref{eq:W_k_inv} follows from the well-known Sherman-Woodbury formula
\citep[Theorem 18.2.8]{Harville:2008wja}.
\end{proof}

@@

Thus far, we have constructed each $\tilde{\bm \Sigma}_k$ in terms of the first
$q$ eigenvectors of $\widehat{\bm \Sigma}$. By utilizing a technique similar to
the so-called kernel trick, we can obtain $\bm U_1$ by computing the
eigendecomposition of a much smaller $N \times N$ matrix when $p \gg N$
\citep{Hastie:2008dt}. Recall that the pooled sample covariance matrix is
$\widehat{\bm \Sigma} = \bm X^{T} \bm X$. Then, because $\bm X = \bm M \bm
D^{1/2} \bm U^{T}$ by the singular value decomposition, we have the
eigendecomposition $\widehat{\bm \Sigma} = \bm U \bm D \bm U^{T}$, where $\bm M
\in \mathbb{R}_{N \times N}$ is orthogonal, $\bm D^{1/2} \in \mathbb{R}_{N
  \times N}^{\ge}$ is a diagonal matrix consisting of the singular values of
$\bm X$, and $\bm U \in \mathbb{R}_{p \times N}$ is orthogonal. Now, $\bm X \bm
X^{T} = \bm M \bm D \bm M^{T}$ is the eigenvalue decomposition of $\bm X \bm
X^{T}$, where $\bm D$ is the matrix of eigenvalues of $\bm X \bm X^{T}$ and the
columns of $\bm M$ are the corresponding eigenvectors. Hence, we obtain $\bm M$
and $\bm D$ from an eigendecomposition of $\bm X \bm X^{T} \in \mathbb{R}_{N
  \times N}$. Next, we compute $\bm U = \bm X^{T} \bm M \bm D^{+/2}$, where $\bm
D^{+/2} = \bm D_q^{-1/2} \oplus \bm 0_{N-q}$. We then determine $q$, the number
of numerically nonzero eigenvalues present in $\bm D$, by calculating the number
of eigenvalues that exceeds some tolerance value, say, $\epsilon = 1 \times
10^{-6}$. After calculating $q$, we extract $\bm U_1$ by retaining only the
first $q$ columns of $\bm U$.

@@

We used version 3.0.0 of the open-source statistical software {\tt R} for the
classification study. We used the implementations of the Witten and Guo
classifiers from the {\tt penalizedLDA} and {\tt rda} packages, respectively, on
CRAN. Because the {\tt rda} package does not perform the authors' ``Min-Min''
rule automatically, we applied this rule in our {\tt R} code. We also utilized
our implementations of the Dudoit, Pang, and Tong classifiers available in the
{\tt sparsediscrim} package. We used the default settings for each classifier
with the exception that we explicitly set prior probabilities as equal, if the
option was available. For the \emph{HDRDA} classifier in
\eqref{eq:hdrda-decomposed}, we examine the classification performance of two
models. For the first \emph{HDRDA} model, we set $\alpha_k = 1$, $k = 1, \ldots,
K$, so that the covariance-matrix estimator \eqref{eq:hdrda-cov} employed
resembled \eqref{eq:ridge-estimator}. We estimated $\lambda$ from a grid of 21
equidistant candidate values between 0 and 1, inclusively. Similarly, we
estimated $\gamma$ from a grid consisting of the values $10^{-1}, \ldots, 10^4$,
and $10^5$. We selected optimal estimates of $\lambda$ and $\gamma$ using
10-fold cross-validation \citep{Hastie:2008dt}. For the second model, we set
$\alpha_k = 1 - \gamma$, $k = 1, \ldots, K$, to resemble
\citeauthor{Friedman:1989tm}'s parameterization. We estimated both $\lambda$ and
$\gamma$ from a grid of 21 equidistant candidate values between 0 and 1,
inclusively. Each of the microarray data sets listed below can be found in
\ref{suppA}.

@@

>>>>

\usepackage{graphicx, amsmath, amssymb, bm, mathtools, amsthm}
\usepackage{natbib, setspace}
\usepackage{algorithm2e}

@@

\begin{proposition}\label{proposition:hdrda-W_k}
Let $\bm W_k$ be defined as above, $k = 1, \ldots, K$. Then, $|\bm W_k| = |
\bm \Gamma_k| |\bm Q_k|$ and
	\begin{align}
		\bm W_k^{-1} &= \bm \Gamma_k^{-1} - \alpha_k(1 - \lambda) \bm \Gamma_k^{-1} \bm U_1^{T} \bm X_k^{T} \bm Q_k^{-1} \bm X_k \bm U_1 \bm \Gamma_k^{-1},\label{eq:W_k_inv}
    \intertext{where}
    \bm Q_k &= \bm I_{n_k} + \alpha_k(1 - \lambda) \bm X_k \bm U_1 \bm \Gamma_k^{-1} \bm U_1^{T} \bm X_k^{T}\label{eq:Q_k}
    \intertext{and}
    \bm \Gamma_k &= \alpha_k \lambda \bm D_q + \gamma \bm I_q\label{eq:Gamma_k}.
	\end{align}
\end{proposition}
\begin{proof}
To calculate $|\bm W_k|$, we apply Theorem 18.1.1 from \cite{Harville:2008wja},
which states that $|\bm A + \bm B \bm T \bm C| = |\bm A| |\bm T| |\bm T^{-1} +
\bm C \bm A^{-1} \bm B|$. Thus, setting $\bm A = \bm \Gamma_k$, $\bm B =
\alpha_k (1 - \lambda) \bm U_1^{T} \bm X_k^{T}$, $\bm T = \bm I_{n_k}$, and $\bm
C = \bm X_k \bm U_1$, we have $|\bm W_k| = |\bm \Gamma_k| |\bm Q_k|$. Also,
\eqref{eq:W_k_inv} follows from the well-known Sherman-Woodbury formula
\citep[Theorem 18.2.8]{Harville:2008wja}.
\end{proof}

@@

Thus far, we have constructed each $\tilde{\bm \Sigma}_k$ in terms of the first
$q$ eigenvectors of $\widehat{\bm \Sigma}$. By utilizing a technique similar to
the so-called kernel trick, we can obtain $\bm U_1$ by computing the
eigendecomposition of a much smaller $N \times N$ matrix when $p \gg N$
\citep{Hastie:2008dt}. Recall that the pooled sample covariance matrix is
$\widehat{\bm \Sigma} = \bm X^{T} \bm X$\footnote{TODO: $1/N$ is missing. Add
  this. How much do things change below?} Then, because $\bm X = \bm M \bm
D^{1/2} \bm U^{T}$ by the singular value decomposition, we have the
eigendecomposition $\widehat{\bm \Sigma} = \bm U \bm D \bm U^{T}$, where $\bm M
\in \mathbb{R}_{N \times N}$ is orthogonal, $\bm D^{1/2} \in \mathbb{R}_{N
  \times N}^{\ge}$ is a diagonal matrix consisting of the singular values of
$\bm X$, and $\bm U \in \mathbb{R}_{p \times N}$ is orthogonal. Now, $\bm X \bm
X^{T} = \bm M \bm D \bm M^{T}$ is the eigenvalue decomposition of $\bm X \bm
X^{T}$, where $\bm D$ is the matrix of eigenvalues of $\bm X \bm X^{T}$ and the
columns of $\bm M$ are the corresponding eigenvectors. Hence, we obtain $\bm M$
and $\bm D$ from an eigendecomposition of $\bm X \bm X^{T} \in \mathbb{R}_{N
  \times N}$. Next, we compute $\bm U = \bm X^{T} \bm M \bm D^{+/2}$, where $\bm
D^{+/2} = \bm D_q^{-1/2} \oplus \bm 0_{N-q}$. We then determine $q$, the number
of numerically nonzero eigenvalues present in $\bm D$, by calculating the number
of eigenvalues that exceeds some tolerance value, say, $\epsilon = 1 \times
10^{-6}$. After calculating $q$, we extract $\bm U_1$ by retaining only the
first $q$ columns of $\bm U$.

\IncMargin{1em}
\begin{algorithm}
  \SetKwData{Left}{left}
  \SetKwData{This}{this}
  \SetKwData{Up}{up}
  \SetKwFunction{Union}{Union}
  \SetKwFunction{FindCompress}{FindCompress}
  \SetKwInOut{Input}{input}
  \SetKwInOut{Output}{output}

  \Input{
    Data matrix $\bm X$\\
    Parameter grid $\{\lambda_g\}_{g=1}^G \times \{\gamma_h\}_{h=1}^H$
  }
  \Output{Optimal Estimates, $\hat{\lambda}$ and $\hat{\gamma}$}
  \BlankLine
  \For{$v \leftarrow 1$ \KwTo $V$ \tcp*[r]{$V$-fold cross-validation}} {
    Partition $v$th fold into $\bm X_{train} \in \mathbb{R}_{N \times p}$ and
    $\bm X_{test} \in \mathbb{R}_{N_T \times p}$

    \For{$k\leftarrow 1$ \KwTo $K$}{
      Extract $\bm X_k \in \mathbb{R}_{n_k \times p}$ from $\bm X_{train}$

      Compute sample mean $\xbar_k$ from $\bm X_k$

      Center $\bm X_k \leftarrow \bm X_k - \bm j_{n_k} \xbar_k^T$    
    }

    $\bm X_c \leftarrow [\bm X_1^T, \ldots, \bm X_K^T]^T$

    Compute $\bm X_c = \bm M_q \bm D_q \bm U_1^T$

    Transform $\bm X_c \leftarrow \bm X_c \bm U_1$

    Transform $\bm X_{test} \leftarrow \bm X_{test} \bm U_1$

    \For{$k \leftarrow 1$ \KwTo $K$} {
      Extract $\bm X_k \in \mathbb{R}_{n_k \times q}$ from $\bm X_c$

      Recompute sample mean $\xbar_k$ from $\bm X_k$
    }

    \For{$(\lambda, \gamma) \in \{\lambda_g\}_{g=1}^G \times \{\gamma_h\}_{h=1}^H$} {
      \For{$k \leftarrow 1$ \KwTo $K$} {      
        Compute $\bm Q_k$ using \eqref{eq:Q_k}

        Compute $\bm \Gamma_k$ using \eqref{eq:Gamma_k}

        Compute $\bm W_k^{-1}$ using \eqref{eq:W_k_inv}

        Compute $|\bm W_k| = |\bm \Gamma_k| |\bm Q_k|$

        Compute $(\bm x - \xbar_k)^{T} \bm U_1 \bm W_k^{-1} \bm U_1^{T} (\bm x - \xbar_k) + \log | \bm W_k |$ \tcp*[r]{TODO: Use matrix form for test observations}
      }
      Classify test observations $\bm X_{test}$ using \eqref{eq:hdrda-decomposed}

      Compute $\#\{\text{Error}_v(\lambda, \gamma)\} \leftarrow$ misclassified test observations
    }   
   }
  Compute $\widehat{\text{Error}}(\lambda, \gamma) = N^{-1} \sum_{v=1}^V \#\{\text{Error}_v(\lambda, \gamma)\}$

  Report optimal $(\hat{\lambda}, \hat{\gamma}) \leftarrow \argmax_{(\lambda, \gamma)} \widehat{\text{Error}}(\lambda, \gamma)$
  \caption{Model selection for the \emph{HDRDA} classifier}
  \label{algo_hdrda_model_selection}
\end{algorithm}
\DecMargin{1em}

The compact SVD is sometimes called the reduced or fast SVD.

Let $\bm j_m \in \mathbb{R}_{m \times 1}$ denote a vector of ones.

@@

We used version 3.0.2 of the open-source statistical software {\tt R} for the
classification study. We used the implementations of the Witten and Guo
classifiers from the {\tt penalizedLDA} and {\tt rda} packages, respectively, on
CRAN. Because the {\tt rda} package does not perform the authors' ``Min-Min''
rule automatically, we applied this rule in our {\tt R} code. We also utilized
our implementations of the Dudoit, Pang, and Tong classifiers available in the
{\tt sparsediscrim} package. We used the default settings for each classifier
with the exception that we explicitly set prior probabilities as equal, if the
option was available. For the \emph{HDRDA} classifier in
\eqref{eq:hdrda-decomposed}, we examine the classification performance of two
models. For the first \emph{HDRDA} model, we set $\alpha_k = 1$, $k = 1, \ldots,
K$, so that the covariance-matrix estimator \eqref{eq:hdrda-cov} employed
resembled \eqref{eq:ridge-estimator}. We estimated $\lambda$ from a grid of 21
equidistant candidate values between 0 and 1, inclusively. Similarly, we
estimated $\gamma$ from a grid consisting of the values $10^{-1}, \ldots, 10^4$,
and $10^5$. We selected optimal estimates of $\lambda$ and $\gamma$ using
10-fold cross-validation \citep{Hastie:2008dt}. For the second model, we set
$\alpha_k = 1 - \gamma$, $k = 1, \ldots, K$, to resemble
\citeauthor{Friedman:1989tm}'s parameterization. We estimated both $\lambda$ and
$\gamma$ from a grid of 21 equidistant candidate values between 0 and 1,
inclusively. Each of the microarray data sets listed below can be found in
\ref{suppA}.

@@

The dimension reduction that we have employed in this paper has reduced the
dimension to the rank$(\widehat{\bm \Sigma})$, $q$. An interesting extension of
our work would reduce the dimension $q$ further to a lower dimension $q_L < q$
using some criterion proposed, perhaps similar to that employed in principal
components analysis. While it is not clear whether classification performance
would improve via such a method, the efficiency of the training and model
selection would certainly improve. Moreover, if $q_L = 2$ or $3$ was selected,
the potential for low-dimensional graphical displays of high-dimensional could
be accomplished.
